Microlensing as a possible probe of event-horizon structure in quasars

### 2016-10-20 paper1/ms.tex
changes
Figure 3 replaced
corrections

Modestly active galactic nucleii are present in the Galactic centre
and in M87.  In these two cases, the central engines ($<0.1\rm\,mas$
on the sky) are close to being resolved through very long baseline
interferometry, which shows preliminary indications of the
jet-launching structures
\citep{2008JPhCS.131a2055D,2012Sci...338..355D,2016ApJ...820...90F}.
Current data do not deliver images but require fitting to predefined
models for the images.  A whole range of models have been applied,
starting from simple geometric models to more complex physical models
\citep{2008Natur.455...78D,2011ApJ...738...38B,2009ApJ...706..497M,2010ApJ...717.1092D}.
The more complicated models nonetheless tend to predict a crescent
shaped silhouette of the black hole.  This motivated
\cite{2013MNRAS.434..765K} to use a simple geometric crescent model to
fit the available sub-mm interferometric data, They argued that the
crescent is nothing but the silhouette of the event horizon.
The great majority of quasars, however, lie at redshifts beyond 2
\citep{2014A&A...563A..54P} and their central engines would be orders
of magnitude smaller on the sky. The direct observations of the black
hole silhouettes of quasars are far beyond foreseeable
instrumentation.

@@

Gravitational microlensing of quasars, reviewed in
Section~\ref{sec:microlensing} below, is a phenomenon seen in quasars
that have been multiply imaged by galaxies.  Individual images in a
multiple-image system can undergo sharp and seemingly random
brightness changes.  This can occur, even for constant intrinic
luminosity, as a consequence of two things: It is a
consequence of two things: the very small size of the central engine,
and granularity of the mass distribution of a lensing galaxy due to
stars.  The latter means that the local magnification is not a smooth
function of source position. It contains a complicated network of
singular curves, known as caustics.
Figure~\ref{fig:magnification_map} shows part of a magnification map
with a few caustics.  The lensed brightness would be given by placing
the source on such a magnification map and integrating the surface
brightness weighted by the magnification.  Most astrophysical sources
straddle several caustics, and hence, their net brightness varies
smoothly with location.  The central engine of quasars, however, is
smaller than the typical spacing between caustics.  As a result, the
lensed brightness undergoes sudden changes as a quasar crosses a
caustic.  Quasar microlensing is mostly observed in the optical
continuum \citep[e.g.,][]{2012A&A...544A..62S} but is also known in
x-rays \citep[e.g.,][]{2016arXiv160207601N} and has been argued for at
cm wavelengths \citep{2000A&A...358..793K}.  The effect supplies an
upper limit on the size of the central engine
\citep[e.g.,][]{2015ApJ...814L..26M} and can be used to study the mass
distribution and kinematics of stars in the lensing galaxy as well
\citep[e.g.,][]{2012ApJ...744..111P}.  Microlensing caustics have an
additional remarkable feature: though they can be very complicated,
they have some universal properties well-known from catastrophe
theory.  In particular, very close to the simplest caustics (known as
folds), the magnification is approximately constant on one side and
$\propto1/\sqrt p$ where $p$ is the transverse distance of the source
from the caustic.  This property will be exploited later.

@@

In Section~\ref{sec:numerics} we carry out source fitting to
lightcurves, with both noise and systematic errors are present.  We
generate a mock lightcurves by taking a crescent source across the
path AB in Figure~\ref{fig:magnification_map}, and then adding noise.
The path simulates crossing a clean but not ideal fold.  We then fit
this lightcurve to templates from both crescent and circular-Gaussian
sources across the path CB.  That is, the templates come from a
similar but not identical caustic, thus deliberately generating a
systematic error.  Despite this systematic error, a crescent source
fits the data, while a Gaussian source is rejected from the $\chi^2$
value.

@@

\begin{figure}
\includegraphics[width = .49\textwidth]{figures/figure_rhalf.eps}
\caption{\label{fig:geom_crescent} Geometry of crescent sources: the source can be seen as a superposition of a smaller dark disk over and inside a larger brighter disk. $R_p$ and $R_n$ mark the radii of the bright and dark disk. $d_c$ and $c$ mark the distance from the center of the bright disk to the centroid and from the dark disk to the centroid.  }
\end{figure}

@@

The previously mentioned abstract parameters can be related to
physical quantities specific to the central region of a quasar.  As
such the luminous region would correspond to the bright accretion disc
that surrounds the black hole. The later's gravity would cause a
shadow in the bright region limited by the extent of the event horizon
of the black hole. Therefore, the radius of the bright disk would
provide an estimate of the size of the accretion disk and the radius
of the dark disc would provide and estimate of the gravitationally
magnified Schwarzschild radius of the black hole $R_{S}^{magnified} =
\Delta t_{dark} \cdot v_p$.  By $R_{S}^{magnified}$ we refer to the
apparent Schwarzschild radius which is larger than the real value at
large distances due to the black hole's own gravity.  Moreover, the
gravitationally magnified value of the Schwarzschild radius is a
monotonic function of the black hole's mass. Therefore, it can be used
to estimate the mass of the black hole if it was not rotating. In the
previous expression, the $\Delta t_{dark}$ denotes the period of
overlap between the black hole shadow and fold caustic.  $v_p$ denotes
the component of the relative velocity of the source and fold which is
perpendicular to the caustic.  The respective velocity is an unknown,
though it can be constrained on a case by case basis to an order of
magnitude or even better. This would require the study of the dynamics
of the stellar structure which contains the gravitational lens.  A
better estimate of the relative velocity would facilitate a better
estimate of the effective non-rotating black hole mass associated with
the black hole shadow.

>>>>

Modestly active galactic nucleii are present in the Galactic centre
and in M87.  In these two cases, the central engines ($<0.1\rm\,mas$
on the sky) are close to being resolved through very long baseline
interferometry, which shows preliminary indications of the
jet-launching structures
\citep{2008JPhCS.131a2055D,2012Sci...338..355D,2016ApJ...820...90F}.
Current data do not deliver images but require fitting to predefined
models for the images.  A whole range of models have been applied,
starting from simple geometric models to more complex physical models
\citep{2008Natur.455...78D,2011ApJ...738...38B,2009ApJ...706..497M,2010ApJ...717.1092D}.
The more complicated models nonetheless tend to predict a crescent
shaped silhouette of the black hole.  This motivated
\cite{2013MNRAS.434..765K} to use a simple geometric crescent model to
fit the available sub-mm interferometric data. They argued that the
crescent is nothing but the silhouette of the event horizon.
The great majority of quasars, however, lie at redshifts beyond 2
\citep{2014A&A...563A..54P} and their central engines would be orders
of magnitude smaller on the sky. The direct observations of the black
hole silhouettes of quasars are far beyond foreseeable
instrumentation.

@@

Gravitational microlensing of quasars, reviewed in
Section~\ref{sec:microlensing} below, is a phenomenon seen in quasars
that have been multiply imaged by galaxies.  Individual images in a
multiple-image system can undergo sharp and seemingly random
brightness changes.  This can occur, even for constant intrinic
luminosity, as a consequence of two things: the very small size of the central engine,
and granularity of the mass distribution of a lensing galaxy due to
stars.  The latter means that the local magnification is not a smooth
function of source position. It contains a complicated network of
singular curves, known as caustics.
Figure~\ref{fig:magnification_map} shows part of a magnification map
with a few caustics.  The lensed brightness would be given by placing
the source on such a magnification map and integrating the surface
brightness weighted by the magnification.  Most astrophysical sources
straddle several caustics, and hence, their net brightness varies
smoothly with location.  The central engine of quasars, however, is
smaller than the typical spacing between caustics.  As a result, the
lensed brightness undergoes sudden changes as a quasar crosses a
caustic.  Quasar microlensing is mostly observed in the optical
continuum \citep[e.g.,][]{2012A&A...544A..62S} but is also known in
X-rays \citep[e.g.,][]{2016arXiv160207601N} and has been argued for at
cm wavelengths \citep{2000A&A...358..793K}.  The effect supplies an
upper limit on the size of the central engine
\citep[e.g.,][]{2015ApJ...814L..26M} and can be used to study the mass
distribution and kinematics of stars in the lensing galaxy as well
\citep[e.g.,][]{2012ApJ...744..111P}.  Microlensing caustics have an
additional remarkable feature: though they can be very complicated,
they have some universal properties well-known from catastrophe
theory.  In particular, very close to the simplest caustics (known as
folds), the magnification is approximately constant on one side and
$\propto1/\sqrt p$ where $p$ is the transverse distance of the source
from the caustic.  This property will be exploited later.

@@

In Section~\ref{sec:numerics} we carry out source fitting to
lightcurves, with both noise and systematic errors are present.  We
generate a mock lightcurves by taking a crescent source across the
path AB in figure~\ref{fig:magnification_map}, and then adding noise.
The path simulates crossing a clean but not ideal fold.  We then fit
this lightcurve to templates from both crescent and circular-Gaussian
sources across the path CB.  That is, the templates come from a
similar but not identical caustic, thus deliberately generating a
systematic error.  Despite this systematic error, a crescent source
fits the data, while a Gaussian source is rejected from the $\chi^2$
value.

@@

\begin{figure}
\includegraphics[width = .49\textwidth]{figures/figure_rhalf2.eps}
\caption{\label{fig:geom_crescent} Geometry of crescent sources: the source can be seen as a superposition of a smaller dark disk over and inside a larger brighter disk. $R_p$ and $R_n$ mark the radii of the bright and dark disk. $d_c$ and $c$ mark the distance from the center of the bright disk to the centroid and from the dark disk to the centroid.  }
\end{figure}

@@

The previously mentioned abstract parameters can be related to
physical quantities specific to the central region of a quasar.  As
such the luminous region would correspond to the bright accretion disc
that surrounds the black hole. The later's gravity would cause a
shadow in the bright region limited by the extent of the event horizon
of the black hole. Therefore, the radius of the bright disk would
provide an estimate of the size of the accretion disk and the radius
of the dark disc would provide and estimate of the gravitationally
magnified Schwarzschild radius of the black hole $R_{S}^{magnified} =
\Delta t_{dark} \cdot v_p$.  By $R_{S}^{magnified}$ we refer to the
apparent Schwarzschild radius which is larger than the real value at
large distances due to the black hole's own gravity.  Moreover, the
gravitationally magnified value of the Schwarzschild radius is a
monotonic function of the black hole's mass. Therefore, it can be used
to estimate the mass of the black hole if it was not rotating. In the
previous expression, the $\Delta t_{dark}$ denotes the period of
overlap between the black hole shadow and fold caustic.  $v_p$ denotes
the component of the relative velocity of the source and fold which is
perpendicular to the caustic.  The respective velocity is an unknown,
though it can be constrained on a case by case basis to an order of
magnitude or even better. This would require the study of the dynamics
of the stellar structure which contains the gravitational lens and the bulk
transverse velocity of the galaxy with respect to the observer.  A
better estimate of the relative velocity would facilitate a better
estimate of the effective non-rotating black hole mass associated with
the black hole shadow.
